Trajectories: how functional traits influence plant growth and shade tolerance across the life-cycle

### 2017-08-09 ms/ms.tex
minor edits, address change

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith NSW 2751, Australia}
\affil[c]{Imperial College, London, United Kingdom}

@@

\authordeclaration{The authors declare that they have no competing financial interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: adaptive.plant@gmail.com}

@@

% 250 words
\begin{abstract}
Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance, extending a widely-used theoretical model that links growth rate in seedlings with a single leaf trait to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of prominent traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Specifically, we demonstrate why for height growth the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results.
\end{abstract}

@@

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the leaf mass per unit leaf area ($\phi$) -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly correlated with relative growth rate in plant mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}, as predicted from a simple mathematical model of growth rate (see below). $\phi$ and leaf lifespan, itself closely correlated with $\phi$, were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, $\phi$ was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. Clearly, traits do not correlate simply and consistently to growth rate.

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations provides the most compelling evidence \citep{Gibert-2016}. That study showed that the strength of the correlation between some traits (including $\phi$ and maximum height) and growth rate was modified by size, while for other traits (including wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, they proposed putative mechanisms to explain their results \citep{Gibert-2016}. In this paper, we test these mechanisms by providing a fully functioning model of plant growth.

\begin{table}[!ht]
\caption{Empirical phenomena explained in this paper.}
\begin{tabular}{p{8cm}}
\toprule
\textbf{Change in growth rate with increasing size (Fig. \ref{fig:conceptual})}\\
 $\,$ Net biomass production: Hump-shaped \citep{Givnish-1988, Koch-2004} \\
 $\,$ Plant mass: Increasing \citep{Sillett-2010, Stephenson-2014} \\
 $\,$ Height: Hump-shaped \citep{Ryan-2006, Sillett-2010, King-2011} \\
 $\,$ Stem-diameter: Hump-shaped \citep{Canham-2004, Canham-2006, Herault-2011} \\
 $\,$ Relative growth rate (all variables): decreasing \citep{Rees-2010, Iida-2014}\\
\textbf{Effect of traits on growth rate (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\omega$: low values $\rightarrow$ smaller seedlings, lower absolute \& higher relative growth \citep{Gibert-2016} \\
 $\,$ $H_{\rm m}$: low values $\rightarrow$ slower growth at at larger sizes \citep{Gibert-2016}\\
 $\,$  $\nu$: low values $\uparrow$ growth irrespective of size, but only in high light \citep{Gibert-2016}\\
 $\,$ $\phi$: low values $\uparrow$ growth when small, not at mid-large sizes \citep{Gibert-2016}\\
 $\,$ $\rho$: low values $\uparrow$ growth, except at largest sizes \citep{Gibert-2016}\\
\textbf{Responsiveness of growth rate to changes in light, $E$ (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\nu$: high values respond more \\
 $\,$ $\phi$: low values respond more \\
 $\,$ $\rho$: low values respond more \citep{Ruger-2012}\\
\textbf{Shade tolerance, \textsc{wplcp} (Fig. \ref{fig:wplcp})}\\
 $\,$ Decreasing with size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}\\
 $\,$ $\nu$: low values $\uparrow$ shade tolerant \citep{Messier-1999, Craine-2005,Baltzer-2007}\\
 $\,$ $\phi$: low values $\downarrow$ shade tolerant \citep{Messier-1999, Poorter-2006, Baltzer-2007, Lusk-2008}$^3$, higher \textsc{lai} \citep{Reich-1992, Gower-1993, Niinemets-2010} \\
 $\,$ $\rho$: low values less shade tolerant \citep{Osunkoya-1996}\\
  \bottomrule
\end{tabular}
\addtabletext{Traits are seed mass ($\omega$), height at maturation ($H_{{\rm mat}}$), leaf nitrogen content per unit leaf area ($\nu$), leaf mass per unit leaf area ($\phi$), and wood density ($\rho$); for details see Tables \ref{tab:definitions}-\ref{tab:traits}. \textsc{lai} = leaf area index. Footnotes: $^1$ Similar responses are predicted for maximum photosynthetic rate per leaf area \& dark respiration rate per leaf area; here both directly related to $\nu$. $^2$ Similar responses are predicted for leaf lifespan; here directly related to $\phi$. $^3$ \citep{Baltzer-2007} reports a strong relationship between \textsc{wplcp} and leaf respiration rate per unit leaf area or mass.}
\label{tab:phenomena}
\end{table}

@@

Interpreting diverse empirical results seeking to link traits to growth rate is challenging because, until recently \citep{Gibert-2016}, we lacked any clear expectations on why the effect of traits may be moderated by size. Generating appropriate expectations is one of the primary roles for theory \citep{Kokko-2007}. A widely-used equation for seedlings suggests that, all else being equal, a seedling's relative growth rate in mass is linearly and negatively related to $\phi$ \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of this equation suggests a similar relationship should hold at larger plant sizes \citep{Enquist-2007}. But as noted above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers expected traits to linearly map onto growth rates, i.e. there is a ``fast'' and ``slow'' growth end for each trait \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One reason theoretical predictions have been lacking or not been supported is that in existing theory the effects of traits is realised mainly via influences on primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as $\phi$ and wood density suggests they influence allocation of biomass among different tissues rather than mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants. On larger plants, growth is measured mainly as increment in either stem diameter or height \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As we outline below, mass production is only one of the factors influencing diameter growth; making theory considering only primary production of limited utility.

@@

\begin{table}[!hb]
 \caption{Variable definitions.}
\centering
\begin{tabular}{p{0.5cm}p{1.75cm}p{5cm}}
\toprule
Symbol & Unit & Description \\
\midrule
\multicolumn{3}{l}{\textbf{Traits (constant through ontogeny)}}\\
$\omega$ & kg & Seed mass  \\
$H_{{\rm mat}}$ & m & Height at maturation\\
$\nu$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf nitrogen per unit leaf area  \\
$\phi$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf mass per area \\
$\rho$ & $\mathrm{kg}\,\mathrm{m}^{-3}$ & Wood density \\
\multicolumn{3}{l}{\textbf{State variables (may vary through ontogeny)}} \\
$H$ & m & Height of a plant\\
$B$ & kg & Biomass originating from parent plant\\
$D$ & m & Stem diameter\\
$M_i$ & kg & Mass of tissue type $i$ retained on plant \\
$A_i$ & m$^2$ & Surface or cross-section are of tissue type $i$\\
$p,\bar{p}$ & mol yr$^{-1}$ m$^{-2}$ & Photosynthetic rate per unit area \\
$r_i$ & mol yr$^{-1}$ kg$^{-1}$  & Respiration rate per unit mass of tissue type $i$ \\
$k_i$ & yr$^{-1}$ & Turnover rate for tissue type $i$ \\
\multicolumn{3}{l}{\textbf{Environmental variables (fixed)}} \\
$E$ & & Canopy openness\\
\multicolumn{3}{l}{\textbf{Other parameters (constant throughout)}} \\
$\alpha_{\rm y}$ &  & Yield, fraction of carbon fixed into biomass\\
$\alpha_{\rm bio}$  & $\mathrm{kg}\,\mathrm{mol}^{-1}$ & Biomass per mol carbon \\
$\eta_c$ & & Crown-shape parameter\\
$\theta$ &  & Sapwood area per unit leaf area\\
$\alpha_{\rm l1}$ & m & Height of plant with leaf area of 1m$^2$ \\
$\alpha_{\rm r1}$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Root mass per unit leaf area \\
$\alpha_{\rm b1}$ &  & Ratio of bark area to sapwood area\\
\bottomrule
\end{tabular}
\addtabletext{For $M_i$, $r_i$, $k_i$ and \(A_i\), subscripts refer to: \({\rm l}\) = leaves, \({\rm b}\) = bark, \({\rm s}\) = sapwood,  \({\rm h}\) = heartwood, \({\rm r}\) = roots, \({\rm st}\) = total stem, \({\rm a}\) =  alive tissue.}
\label{tab:definitions}
\end{table}

@@

Earlier studies on seedlings related relative growth rate in mass to $\phi$ \citep{Blackman-1919, Lambers-1992, Cornelissen-1996, Wright-2000}. This model can be derived from eqn \ref{eq:dbdt}, as a special case of the more-extended model described in the following sections, as follows. Ignoring all turnover terms as well as the respiration terms for non-leaf tissues in eqn \ref{eq:dbdt}, net production becomes a linear function of leaf area, making relative growth rate in mass:
\begin{equation}\label{eq:RGR}
\frac{{\rm d}B}{{\rm d}t}\frac{1}{M_{\rm a}}  \approx P_{\rm net} \times \phi^{-1} \times \frac{M_{\rm l}}{M_{\rm a}}, \end{equation}
where $P_{\rm net} = \alpha_{\rm bio} \, \alpha_{\rm y}\,(\bar{p}(E) - r_{\rm l})$. Although eqn \ref{eq:RGR} captures patterns of growth in seedlings in relation to $\phi$ \citep{Wright-2000}, this approximation does not easily extend to the variables that are routinely collected for large trees such as height or stem diameter. The derivations below clarify these links.

@@

We compared the assumptions outlined in Table \ref{tab:allometry}a to data sourced from the \textsc{baad} \citep{Falster-2015b}. We also evaluated an additional prediction arising from the eqns in Table \ref{tab:allometry}a. The amount of live stem tissue supporting each unit of leaf area is predicted to increase linearly with height as
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Note that $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are all traits and so assumed approximately constant for any given species.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][!ht]
\centering
\includegraphics[width=11.4cm,height=11.4cm, keepaspectratio]{main/growth_light_height.pdf}
\caption{\textbf{Effect of four traits on height growth rate for different-sized plants.}
Growth rates were simulated using the {\plant} model, applying the trade-offs describing in Table \ref{tab:traits}. Each panel shows how growth is influenced by a different trait for plants of a given height, and across a series of canopy openness values from completely open (light blue,  $E=1$) to heavily shaded (dark line, $E=0.25$). For any given value of trait and $E$, plants were grown to the desired height and their growth rate estimated. The white regions indicate trait ranges that are typically observed in real systems. Figs. \ref{fig:growth_light_dia}-\ref{fig:growth_light_mass} show similar plots but with growth measured as stem diameter, stem area, or plant mass. Changes in trait-growth relationships are summarised in Table \ref{tab:responses}.
\label{fig:growth_light_height}}
\end{SCfigure*}

@@

We analysed the response of growth rate to five different traits under the assumed trade-offs (Table \ref{tab:traits}). We considered changes in absolute and relative growth rates for mass, height, stem area and stem diameter. The first two traits considered modify behaviour primarily at the start and end of an individual's growth trajectory and are therefore termed ``ontogenetic traits''. The remaining three traits are termed ``development traits'' because they moderate the rate of movement along an individual's growth trajectory. Across the five different traits, we observed four relatively distinct types of response. These responses are summarised in Table \ref{tab:responses} and described in more detail below.

@@

\subsubsection{Development traits}  The response of growth rate to changes in $\nu$ is relatively straightforward: there is an optimum value of $\nu$ that maximises height growth rate in a given light environment $E$ and does not vary with height (Fig. \ref{fig:growth_light_height}). As $E$ increases, the optimal $\nu$ also increases. The invariance of the growth-trait relationship with respect to size arises as follows. The direct physiological effect of $\nu$ is to increase the maximum potential photosynthetic rate of leaves, with a cost of higher respiration rate. Both the cost and benefits of $\nu$ appear within $\frac{ {\rm d}B} {{\rm d}t}$, implying the direction of correlation between trait and growth rate depends crucially on the change in mass production per $\nu$.
From eqn \ref{eq:dbdt}, $\frac{\partial \left( \frac{ {\rm d}B} {{\rm d}t}\right)}{\partial nu} = A_{\rm l} \, (\frac{ \partial \bar{p}(E)}{\partial nu}  - \frac{ \partial r_{\rm l}}{\partial nu} )$. As both $\bar{p}(E)$ and $r_{\rm l}$ are expressed per unit area and independent of height, the optimal value is also independent of height.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][b]
\centering
\includegraphics[width=11.4cm,keepaspectratio]{main/lcp_trait.pdf}
\caption{\textbf{Effect of three development traits on shade tolerance.}
Panels show effect of traits on level of canopy openness that causes net production (eqn \ref{eq:dbdt}) to be zero. Different lines indicate relationship for plants with specified height, from short (light blue,  $H=0.5$m) to tall (dark line, $H=20$m). The white regions indicate trait ranges that are typically observed in real systems.
\label{fig:wplcp}}
\end{SCfigure*}

@@

The model presented here extends a widely-used theoretical model for seedlings, which links mass-based growth rate to the trait $\phi$ \citep{Lambers-1992, Wright-2000}, to larger plants and other traits. Importantly, the seedling model can be derived as a special case of the new-extended model (eqn \ref{eq:RGR}). Unlike the original, the extended model also predicts a relationship between $\phi$ and growth rate that changes with size. In particular, the correlation shifts from being strongly negative in seedlings to being absent, or even possibly positive in larger plants (Fig. \ref{fig:growth_light_height}), irrespective of whether growth rate is estimate via height, stem diameter, stem area or total mass. This shift, which matches empirical evidence \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015, Gibert-2016}, occurs because the benefits of cheap leaf deployment diminish with plant size. As seedlings, leaves comprise a large part of the plant (Fig. \ref{fig:conceptual}a). Decreasing $\phi$ then has an overwhelmingly positive effect on growth rate because the effect of increasing $\frac{\ud A_{\rm l}}{\ud M_{\rm l}}$ is large compared to the other terms in eqn \ref{eq:daldmt}. As plants increase in size, however, the amount of supporting tissue increases (Fig \ref{fig:assumptions}d), decreasing the benefit of cheap leaf construction. Consequently, the effect of $\phi$ on leaf turnover comes to dominate at larger sizes, and as such, the effect of $\phi$ on height, diameter and, mass growth shifts from negative to either flat or mildly positive  (Table \ref{tab:responses}).

@@

\matmethods{
\subsection{Trait-based trade-offs}
The trade-offs described in Table \ref{tab:traits} were embedded within the growth model as follows.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Evolution \& Ecology Research Centre, School of Biological, Earth and
Environmental Sciences, University of New South Wales, Sydney, NSW 2052, Australia}
\affil[b]{Hawkesbury Institute for the Environment, Western Sydney University, Locked Bag 1797, Penrith NSW 2751, Australia}
\affil[c]{Imperial College, London, United Kingdom}

@@

\authordeclaration{The authors declare that they have no competing financial interests.\\

Code reproducing the results is available at \href{https://github.com/traitecoevo/growth_trajectories}{github.com/traitecoevo/growth\_trajectories}.
}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: daniel.falster@unsw.edu.au}

@@

% 250 words
\begin{abstract}
Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance by extending a widely-used model, linking growth rate in seedlings with a single leaf trait, to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Specifically, we demonstrate why for height growth the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results.
\end{abstract}

@@

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area ($\phi$) -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly correlated with relative growth rate in plant mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}, as predicted from a simple mathematical model of growth rate (see below). $\phi$ and leaf lifespan, itself closely correlated with $\phi$, were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, $\phi$ was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. Clearly, traits do not correlate simply and consistently to growth rate.

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations provides the most compelling evidence \citep{Gibert-2016}. That study showed that the strength of the correlation between some traits (including $\phi$ and maximum height) and growth rate was modified by size, while for other traits (including wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Here, we extend previous work \citep{Gibert-2016} to provide a mechanistic explanation of these empirical patterns via a model of plant growth and allocation.

\begin{table}[!ht]
\caption{Empirical phenomena explained in this paper.}
\begin{tabular}{p{8cm}}
\toprule
\textbf{Change in growth rate with increasing size (Fig. \ref{fig:conceptual})}\\
 $\,$ Net biomass production: Hump-shaped \citep{Givnish-1988, Koch-2004} \\
 $\,$ Plant mass: Increasing \citep{Sillett-2010, Stephenson-2014} \\
 $\,$ Height: Hump-shaped \citep{Ryan-2006, Sillett-2010, King-2011} \\
 $\,$ Stem-diameter: Hump-shaped \citep{Canham-2004, Canham-2006, Herault-2011} \\
 $\,$ Relative growth rate (all variables): decreasing \citep{Rees-2010, Iida-2014}\\
\textbf{Effect of traits on growth rate (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\omega$: low values $\rightarrow$ smaller seedlings, lower absolute \& higher relative growth \citep{Gibert-2016} \\
 $\,$ $H_{\rm m}$: low values $\rightarrow$ slower growth at at larger sizes \citep{Gibert-2016}\\
 $\,$  $\nu$: low values $\uparrow$ growth irrespective of size, but only in high light \citep{Gibert-2016}\\
 $\,$ $\phi$: low values $\uparrow$ growth when small, not at mid-large sizes \citep{Gibert-2016}\\
 $\,$ $\rho$: low values $\uparrow$ growth, except at largest sizes \citep{Gibert-2016}\\
\textbf{Responsiveness of growth rate to changes in light, $E$ (Fig. \ref{fig:growth_light_height})}\\
 $\,$ $\nu$: high values respond more \\
 $\,$ $\phi$: low values respond more \\
 $\,$ $\rho$: low values respond more \citep{Ruger-2012}\\
\textbf{Shade tolerance, \textsc{wplcp} (Fig. \ref{fig:wplcp})}\\
 $\,$ Decreasing with size \citep{Givnish-1988, Kneeshaw-2006, Lusk-2008}\\
 $\,$ $\nu$: low values $\uparrow$ shade tolerant \citep{Messier-1999, Craine-2005,Baltzer-2007}\\
 $\,$ $\phi$: low values $\downarrow$ shade tolerant \citep{Messier-1999, Poorter-2006, Baltzer-2007, Lusk-2008}$^3$, higher \textsc{lai} \citep{Reich-1992, Gower-1993, Niinemets-2010} \\
 $\,$ $\rho$: low values less shade tolerant \citep{Osunkoya-1996}\\
  \bottomrule
\end{tabular}
\addtabletext{Traits are seed mass ($\omega$), height at maturation ($H_{{\rm mat}}$), leaf nitrogen content per unit leaf area ($\nu$), leaf mass per unit leaf area ($\phi$), and wood density ($\rho$); for details see Tables \ref{tab:definitions}-\ref{tab:traits}. \textsc{lai} = leaf area index. Footnotes: $^1$ Similar responses are predicted for maximum photosynthetic rate per leaf area \& dark respiration rate per leaf area; here both directly related to $\nu$. $^2$ Similar responses are predicted for leaf lifespan; here directly related to $\phi$. $^3$ \citep{Baltzer-2007} reports a strong relationship between \textsc{wplcp} and leaf respiration rate per unit leaf area or mass.

}
\label{tab:phenomena}
\end{table}

@@

Interpreting diverse empirical results seeking to link traits to growth rate is challenging because, until recently \citep{Gibert-2016}, we lacked any clear expectations on why the effect of traits may be moderated by size. A widely-used equation for seedlings suggests that, all else being equal, a seedling's relative growth rate in mass is linearly and negatively related to $\phi$ \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of this equation suggests a similar relationship should hold at larger plant sizes \citep{Enquist-2007}. But as noted above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers have expected traits to map directy to growth rates, i.e. there is a ``fast'' and ``slow'' growth traits \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One reason theoretical predictions have been lacking or not been supported is that in existing theory the effects of traits is realised mainly via influences on primary productivity (photosynthesis - respiration) \citep{Wright-2000, Enquist-2007}. By contrast, the physiology of traits such as $\phi$ and wood density suggests they influence allocation of biomass among different tissues rather than mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants. On larger plants, growth is measured mainly as increment in either stem diameter or height \citep{Purves-2008, Anderson-2015, Kunstler-2016}. As we outline below, height and diameter growth depend not only on mass production but also allocation of biomass and tissue construction costs, meaning theory considering only primary production of limited utility.

@@

\begin{table}[!hb]
 \caption{Variable definitions.}
\centering
\begin{tabular}{p{0.5cm}p{1.75cm}p{5cm}}
\toprule
Symbol & Unit & Description \\
\midrule
\multicolumn{3}{l}{\textbf{Traits (constant through ontogeny)}}\\
$\omega$ & kg & Seed mass  \\
$H_{{\rm mat}}$ & m & Height at maturation\\
$\nu$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf nitrogen per unit leaf area  \\
$\phi$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Leaf mass per area \\
$\rho$ & $\mathrm{kg}\,\mathrm{m}^{-3}$ & Wood density \\
\multicolumn{3}{l}{\textbf{State variables (may vary through ontogeny)}} \\
$H$ & m & Height of a plant\\
$B$ & kg & Biomass originating from parent plant\\
$D$ & m & Stem diameter\\
$M_i$ & kg & Mass of tissue type $i$ retained on plant \\
$A_i$ & m$^2$ & Surface or cross-section are of tissue type $i$\\
$p,\bar{p}$ & mol yr$^{-1}$ m$^{-2}$ & Photosynthetic rate per unit area \\
$r_i$ & mol yr$^{-1}$ kg$^{-1}$  & Respiration rate per unit mass of tissue type $i$ \\
$k_i$ & yr$^{-1}$ & Turnover rate for tissue type $i$ \\
\multicolumn{3}{l}{\textbf{Environmental variables (fixed)}} \\
$E$ & & Canopy openness\\
\multicolumn{3}{l}{\textbf{Other parameters (constant throughout)}} \\
$\alpha_{\rm y}$ &  & Yield, fraction of carbon fixed into biomass\\
$\alpha_{\rm bio}$  & $\mathrm{kg}\,\mathrm{mol}^{-1}$ & Biomass per mol carbon \\
$\eta_c$ & & Crown-shape parameter\\
$\theta$ &  & Sapwood area per unit leaf area\\
$\alpha_{\rm l1}$ & m & Height of plant with leaf area of 1m$^2$ \\
$\alpha_{\rm r1}$ & $\mathrm{kg}\,\mathrm{m}^{-2}$ & Root mass per unit leaf area \\
$\alpha_{\rm b1}$ &  & Ratio of bark area to sapwood area\\
\bottomrule
\end{tabular}
\addtabletext{For $M_i$, $r_i$, $k_i$ and \(A_i\), subscripts refer to: \({\rm l}\) = leaves, \({\rm b}\) = bark, \({\rm s}\) = sapwood,  \({\rm h}\) = heartwood, \({\rm r}\) = roots, \({\rm st}\) = total stem, \({\rm a}\) =  alive tissue.

}
\label{tab:definitions}
\end{table}

@@

Earlier studies on seedlings related relative growth rate in mass to $\phi$ \citep{Blackman-1919, Lambers-1992, Cornelissen-1996, Wright-2000}. This model can be derived from eqn \ref{eq:dbdt} as follows and is a special case of the more-extended model described below. Ignoring all turnover terms as well as the respiration terms for non-leaf tissues in eqn \ref{eq:dbdt}, net production becomes a linear function of leaf area, making relative growth rate in mass:
\begin{equation}\label{eq:RGR}
\frac{{\rm d}B}{{\rm d}t}\frac{1}{M_{\rm a}}  \approx P_{\rm net} \times \phi^{-1} \times \frac{M_{\rm l}}{M_{\rm a}}, \end{equation}
where $P_{\rm net} = \alpha_{\rm bio} \, \alpha_{\rm y}\,(\bar{p}(E) - r_{\rm l})$. Although eqn \ref{eq:RGR} captures patterns of growth in seedlings in relation to $\phi$ \citep{Wright-2000}, this approximation does not easily extend to the variables that are routinely collected for large trees such as height or stem diameter. The derivations below clarify these links.

@@

We compared the assumptions outlined in Table \ref{tab:allometry}a to data sourced from the Biomass and Allometry Database \citep[\textsc{baad}][]{Falster-2015b}. We also evaluated an additional prediction arising from the eqns in Table \ref{tab:allometry}a. The amount of live stem tissue supporting each unit of leaf area is predicted to increase linearly with height as
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Note that $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are all traits and so assumed approximately constant for any given species.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][!ht]
\centering
\includegraphics[width=12.4cm,height=12.4cm, keepaspectratio]{main/growth_light_height.pdf}
\caption{\textbf{Effect of four traits on height growth rate for different-sized plants.}
Growth rates were simulated using the {\plant} model, applying the trade-offs describing in Table \ref{tab:traits}. Each panel shows how growth is influenced by a different trait for plants of a given height, and across a series of canopy openness values from completely open (light blue,  $E=1$) to heavily shaded (dark line, $E=0.25$). For any given value of trait and $E$, plants were grown to the desired height and their growth rate estimated. The white regions indicate trait ranges that are typically observed in real systems. Figs. \ref{fig:growth_light_dia}-\ref{fig:growth_light_mass} show similar plots but with growth measured as stem diameter, stem area, or plant mass. Changes in trait-growth relationships are summarised in Table \ref{tab:responses}.
\label{fig:growth_light_height}}
\end{SCfigure*}

@@

We analysed the response of growth rate to five different traits under the assumed trade-offs (Table \ref{tab:traits}). We considered changes in absolute and relative growth rates for mass, height, stem area and stem diameter. The first two traits considered modify behaviour primarily at the start and end of an individual's growth trajectory and are therefore termed ``ontogenetic traits''. The remaining three traits are termed ``developmental traits'' because they moderate the rate of movement along an individual's growth trajectory. Across the five different traits, we observed four relatively distinct types of response. These responses are summarised in Table \ref{tab:responses} and described in more detail below.

@@

\subsubsection{Developmental traits}  The response of growth rate to changes in $\nu$ is relatively straightforward: there is an optimum value of $\nu$ that maximises height growth rate in a given light environment $E$ and does not vary with height (Fig. \ref{fig:growth_light_height}). As $E$ increases, the optimal $\nu$ also increases. The invariance of the growth-trait relationship with respect to size arises as follows. The direct physiological effect of $\nu$ is to increase the maximum potential photosynthetic rate of leaves, with a cost of higher respiration rate. Both the cost and benefits of $\nu$ appear within $\frac{ {\rm d}B} {{\rm d}t}$, implying the direction of correlation between trait and growth rate depends crucially on the change in mass production per $\nu$.
From eqn \ref{eq:dbdt}, $\frac{\partial \left( \frac{ {\rm d}B} {{\rm d}t}\right)}{\partial nu} = A_{\rm l} \, (\frac{ \partial \bar{p}(E)}{\partial nu}  - \frac{ \partial r_{\rm l}}{\partial nu} )$. As both $\bar{p}(E)$ and $r_{\rm l}$ are expressed per unit area and independent of height, the optimal value is also independent of height.

@@

\begin{SCfigure*}[\sidecaptionrelwidth][b]
\centering
\includegraphics[width=11.4cm,keepaspectratio]{main/lcp_trait.pdf}
\caption{\textbf{Effect of three developmental traits on shade tolerance.}
Panels show effect of traits on level of canopy openness that causes net production (eqn \ref{eq:dbdt}) to be zero. Different lines indicate relationship for plants with specified height, from short (light blue,  $H=0.5$m) to tall (dark line, $H=20$m). The white regions indicate trait ranges that are typically observed in real systems.
\label{fig:wplcp}}
\end{SCfigure*}

@@

The model presented here extends a widely-used model for seedlings, linking mass-based growth rate to the trait $\phi$ \citep{Lambers-1992, Wright-2000}, to larger plants and other traits. Importantly, the seedling model can be derived as a special case of the new-extended model (eqn \ref{eq:RGR}). Unlike the original, the extended model also predicts a relationship between $\phi$ and growth rate that changes with size. In particular, the correlation shifts from being strongly negative in seedlings to being absent, or even possibly positive in larger plants (Fig. \ref{fig:growth_light_height}), irrespective of whether growth rate is estimate via height, stem diameter, stem area or total mass. This shift, which matches empirical evidence \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015, Gibert-2016}, occurs because the benefits of cheap leaf deployment diminish with plant size. As seedlings, leaves comprise a large part of the plant (Fig. \ref{fig:conceptual}a). Decreasing $\phi$ then has an overwhelmingly positive effect on growth rate because the effect of increasing $\frac{\ud A_{\rm l}}{\ud M_{\rm l}}$ is large compared to the other terms in eqn \ref{eq:daldmt}. As plants increase in size, however, the amount of supporting tissue increases (Fig \ref{fig:assumptions}d), decreasing the benefit of cheap leaf construction. Consequently, the effect of $\phi$ on leaf turnover comes to dominate at larger sizes, and as such, the effect of $\phi$ on height, diameter and, mass growth shifts from negative to either flat or mildly positive  (Table \ref{tab:responses}).

@@

\matmethods{
\vspace{-1em}
\subsection{Trait-based trade-offs}
The trade-offs described in Table \ref{tab:traits} were embedded within the growth model as follows.
